A simplified Plackett-Luce likelihood function for rank orders

### 2024-03-25 inst/hankin-ejs.tex
rewording to include actual event

One long-standing contention in the athletic world is that of lane
bias~\cite{munro2022}.  Running in the inner lane is frequently held
to be a disadvantage; mechanisms include biomechanical issues arising
from centrifugal acceleration, to competitors' inability to see their
rivals.  Here attention is restricted to heats, for which lane placing
is random.  I consider every heat in Olympic running from 2004-2020 (a
total of 41 heats, each of 8 or 9 competitors); only competitors'
ranks are used.  Figure~\ref{plotolymp} shows a log-likelihood curve
for the Bradley-Terry strength of the focal competitor, here the
runner in the innermost lane.  We see that the evaluate of $0.574$ is
very close to the neutral value of $0.5$, for which the support is
only $0.16$, far short of Edwards's two units of support criterion for
rejection~\cite{edwards1972}.  This dataset contains no evidence to
support lane bias.

@@

(that is, a logistic regression of generalized Bradley-Terry strength
$a$ against time, with $\alpha$ and $\beta$ being linear
coefficients).  Contours of equal support are shown in
Figure~\ref{showchecolike}, again normalized so that
$\mathcal{S}(\hat{\alpha},\hat{\beta})=0$.  We see the expected
elliptical contours, and further $H_0\colon\beta=0$ may be rejected in
favour of $\alpha=-0.27$, $\beta=0.081$, on the grounds that
$\mathcal{S}(\alpha,0)\leqslant 9.2$.  Alternatively we may observe
that $2\mathcal{S}=18.4$ would correspond to a $p$-value of about
$1.7\times 10^{-5}$ on its asymptotic $\chi^2_1$ null distribution.

>>>>

One long-standing contention in the athletic world is that of lane
bias~\cite{munro2022}.  Running in the inner lane is frequently held
to be a disadvantage; mechanisms include biomechanical issues arising
from centrifugal acceleration, to competitors' inability to see their
rivals.  Here attention is restricted to heats, for which lane placing
is random.  I consider every heat in Men's 100m Olympic running from
2004-2020 (a total of 41 heats, each of 8 or 9 competitors); only
competitors' ranks are used.  Figure~\ref{plotolymp} shows a
log-likelihood curve for the Bradley-Terry strength of the focal
competitor, here the runner in the innermost lane.  We see that the
evaluate of $0.574$ is very close to the neutral value of $0.5$, for
which the support is only $0.16$, far short of Edwards's two units of
support criterion for rejection~\cite{edwards1972}.  This dataset
contains no evidence to support lane bias.

@@

(that is, a logistic regression of generalized Bradley-Terry strength
$a$ against time $t$, with $\alpha$ and $\beta$ being linear
coefficients).  Contours of equal support are shown in
Figure~\ref{showchecolike}, again normalized so that
$\mathcal{S}(\hat{\alpha},\hat{\beta})=0$.  We see the expected
elliptical contours, and further $H_0\colon\beta=0$ may be rejected in
favour of $\alpha=-0.27$, $\beta=0.081$, on the grounds that
$\mathcal{S}(\alpha,0)\leqslant 9.2$.  Alternatively we may observe
that $2\mathcal{S}=18.4$ would correspond to a $p$-value of about
$1.7\times 10^{-5}$ on its asymptotic $\chi^2_1$ null distribution.
